---
title: "MABM Analysis and Power Analysis Overview"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This file is meant to document the general process of preparing and analyzing the MABM data, as well as using the results of this analysis to conduct a power analysis exploring how survey design influences our ability to detect trends in bat populations using the MABM protocol.

## Install and load necessary R packages

```{r}
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
if (!requireNamespace("nrsmisc", quietly = TRUE)) devtools::install_github("adamdsmith/nrsmisc")
if (!requireNamespace("MABMreportr", quietly = TRUE)) devtools::install_github("adamdsmith/MABMreportr")
pacman::p_load(nrsmisc, raster, dplyr, sf, lubridate, maptools)
other_packs <- c("riem", "geonames", "geosphere", "readxl", "rlang")
invisible(sapply(other_packs, function(i) {
    if (!requireNamespace(i, quietly = TRUE))
        install.packages(i, verbose = FALSE)
}))
options(stringsAsFactors = FALSE, scipen = 999)
```


## Export relevant tables from MABM Access database

```{r}
# You may need to adjust the path to the MABM database
MABMreportr:::setup_MABM_reports(MABM_access_db = "../Data/R4_MABM_database2.accdb",
                                 export_dir = normalizePath("./Output/Raw_DB_output"))
```

## Create survey details

```{r}
MABM_routes <- readxl::read_excel("./Output/Raw_DB_output/MABM_routes.xlsx") %>%
  select(station = ORGNAME,
         site = Site_Name,
         site_notes = Loc_Notes,
         len_mi = Rte_length,
         lat = Centroid_Y_Coord,
         lon = Centroid_X_Coord,
         state = State) %>%
  filter(!is.na(lat)|!is.na(lon)) %>%
  rowwise() %>%
  mutate(wx_stns = paste(find_wx_stns(lon, lat, asos_only = TRUE), collapse = ", ")) %>%
  arrange(site)

# Get timezone of each station
options(geonamesHost = "api.geonames.org")
options(geonamesUsername = "adam_d_smith")
timez <- character(nrow(MABM_routes))
timez <- pbapply::pbsapply(seq_len(nrow(MABM_routes)), function(i) {
    timez[i] <- as.character(geonames::GNtimezone(MABM_routes$lat[i], MABM_routes$lon[i])$timezoneId)
})
MABM_routes$timezone <- timez
# Set timezone for Patoka River Lake manually as they operate on Central time
MABM_routes[grep("PtkNWR", MABM_routes$site), "timezone"] <- "America/Chicago"
timez <- unique(MABM_routes$timezone)
saveRDS(MABM_routes, file = "./Output/MABM_routes.rds")

# MABM survey details
MABM_survey_info <- readxl::read_excel("./Output/Raw_DB_output/MABM_survey_details.xlsx") %>%
  select(site = starts_with("Site Name"),
         surv_date = starts_with("Date Start"),
         surv_time = starts_with("Time Start"),
         gps = starts_with("GPS"),
         complete = starts_with("Rt Compl"),
         notes = Notes) %>%
  left_join(MABM_routes[, c("site", "timezone")], by = "site") %>%
  rowwise() %>%
  mutate(surv_dt_gmt = ymd_hm(paste(surv_date, surv_time), tz = timezone)) %>%
  ungroup() %>%
  mutate(gps = as.logical(gps),
         complete = as.logical(complete),
         notes = ifelse(is.na(notes), "", notes)) %>%
  select(site, surv_date, surv_dt_gmt, gps, complete, notes)
attributes(MABM_survey_info$surv_dt_gmt)$tzone <- "GMT"    
```

## Get weather near sunset on day of survey

```{r}
## Get sunset times to calculate time since set and associated weather data
sunsets <- left_join(MABM_survey_info[, c("site", "surv_date")],
                     MABM_routes[, c("site", "timezone", "lat", "lon")],
                     by = "site") %>%
  left_join(MABM_routes[, c("site", "wx_stns")], by = "site") %>%
  rowwise() %>%
  mutate(ss_dt_gmt = get_sun(lon, lat, surv_date, direction = "sunset")$sunset) %>%
  ungroup()

# Import function to get wx near sunset
source("./R/get_sunset_wx.R")

wx <- sunsets %>%
  group_by(site, surv_date, ss_dt_gmt) %>%
  # Add weather data at/near sunset
  # ss_temp = temperature (Celsius) near sunset
  # ss_wsp = wind speed (m/s) near sunset
  # ss_wx_t_diff = time difference (min) between sunset and weather observations
  # wx_stn = weather station of observations
  # Takes ~ 2 h
  do(get_sunset_wx(.$wx_stns, .$surv_date, .$ss_dt_gmt, .$lat, .$lon)) %>%
  ungroup()
```

## Combine weather and survey information
```{r}
MABM_survey_info <- MABM_survey_info %>%
  left_join(sunsets, by = c("site", "surv_date")) %>%
  select(-wx_stns) %>%
  left_join(wx, by = c("site", "surv_date", "ss_dt_gmt")) %>%
  mutate(ss_surv_t_diff = as.numeric(round(difftime(surv_dt_gmt, ss_dt_gmt, units = c("mins")), 2))) %>%
  filter(yday(surv_date) %in% 145:213, # 25 May through 31 July
         year(surv_date) %in% 2012:2017) %>%
  mutate(doy = yday(surv_date))
saveRDS(MABM_survey_info, file = "./Output/MABM_survey_info.rds")
```


# Retrieve raw call data

```{r}
MABM_calls_raw <- readxl::read_excel("./Output/Raw_DB_output/MABM_calls.xlsx") %>%
  select(site = Site_Name, lat = LAT, lon = LONG, FILENAME, spp = A_SP, A_Spper,
         A_Dprob, n_pulse = A_pulse, surv_date = contains("Date Start")) %>%
  mutate(spp_pulse_prop = round(A_Spper, 1),
         spp_dprob = round(A_Dprob, 2),
         year = as.integer(format(surv_date, "%Y"))) %>%
  left_join(MABM_routes[, c("site", "timezone")], by = "site")

fixed_tz <- lapply(timez, function(t) {
  tmp <- MABM_calls_raw %>%
    filter(timezone == t) %>%
      mutate(call_date = ifelse(substr(FILENAME, 5, 6) %in% c("24", "01", "02", "03", "04"),
                                as.character(surv_date + as.difftime(1, units = "days")),
                                as.character(surv_date)),
             call_dt = ymd_hms(paste(call_date,
                                     paste(ifelse(substr(FILENAME, 5, 6) != "24",
                                                  substr(FILENAME, 5, 6), "00"),
                                           substr(FILENAME, 7, 8),
                                           substr(FILENAME, 10, 11), sep = ":")),
                               tz = t),
             call_dt_gmt = call_dt)
  attributes(tmp$call_dt_gmt)$tzone <- "GMT"
  tmp
})

MABM_calls_raw <- do.call("rbind", fixed_tz) %>%
  left_join(sunsets[c("site", "surv_date", "ss_dt_gmt")], by = c("site", "surv_date")) %>%
  # Time of each call relative to sunset
  mutate(ss_call_t_diff = as.numeric(round(difftime(call_dt_gmt, ss_dt_gmt, units = c("mins")), 2))) %>%
  select(site, year, surv_date, ss_dt_gmt, spp, spp_pulse_prop, spp_dprob, n_pulse, 
         call_lat = lat, call_lon = lon, call_dt, ss_call_t_diff) %>%
  arrange(site, call_dt)
saveRDS(MABM_calls_raw, file = "./Output/MABM_calls_raw.rds")
```

## Species code/common name lookup table

```{r}
bat_spp_info <- readxl::read_excel("./Output/Raw_DB_output/MABM_spp_details.xlsx") %>%
  select(spp = A_SP, spp_cn = CommonName)
saveRDS(bat_spp_info, file = "./Output/bat_spp_info.rds")
```

## Species filter matrix

```{r}
spp_filter_matrix <- read.csv("../Data/spp_filter_matrix.csv", stringsAsFactors = FALSE) %>%
    select(site = Site_Name, everything(), -ID, -ORGNAME)
saveRDS(spp_filter_matrix, file = "./Output/spp_filter_matrix.rds")
```

## Create land cover based data set
```{r}
# This takes a while. Progress bar is shown...
source("./R/extract_route_NLCD.R")
```

## Generate data set of species counts by survey
```{r}
# Get all relevant survey dates and sites for completing call data
all_survs <- unique(MABM_survey_info[, c("site", "surv_date")])

# Bat species of interest
boi <- c("LABO", "EPFU", "NYHU", "PESU", "MYLU")

expand.grid.df <- function(...) Reduce(function(...) merge(..., by=NULL), list(...))

MABM_data <- lapply(boi, function(i) {
    i_n <- as.name(i)
    valid_sites <- spp_filter_matrix %>%
        filter(UQ(i_n) == 1) %>% pull(site)
    valid_survs <- filter(all_survs, site %in% valid_sites)

    boi_calls <- MABM_calls_raw %>%
        filter(spp == i) %>%
        group_by(site, surv_date, spp) %>%
        summarise(count = n())
    
    out <- expand.grid.df(valid_survs, spp = data.frame(spp = i)) %>%
        left_join(boi_calls, c("site", "surv_date", "spp")) %>%
        mutate(count = ifelse(is.na(count), 0, count),
               year = year(surv_date)) %>%
        arrange(site, surv_date)
})
names(MABM_data) <- boi
saveRDS(MABM_data, file = "./Output/MABM_spp_counts.rds")
```

## Fit generalized linear mixed models of species counts

We fit relative abundance GLMMs for the five most-detected bat species, including:

- Eastern red bat (*Lasiurus borealis*; LABO)
- Big brown bat (*Eptesicus fuscus*; EPFU)
- Evening bat (*Nycticeius humeralis*; NYHU)
- Eastern pipistrelle (*Perimyotis subflavus*; PESU)
- Little brown myotis (*Myotis lucifugus*; MYLU)

The species-specific GLMMs were constructed with the following line of code. Look within `R/glmmTMB_fits.R` for details.

```{r}
source("./R/glmmTMB_fits.R")
```

## Power analysis

Parameter estimates from the above species-specific GLMMs informed the power analysis defined and conducted in the following line of code. See `Output/power_analysis_setup.pdf` for a summary of the power analysis parameters. Look within `R/power_analysis.R` for details.

```{r}
source("./R/power_analysis.R")
```
